\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=1in}

\title{ValhallaBase: Structure-First Inference with\\Auditable Routing, Algebraic Output Gating,\\and Context-Efficient Session Memory}
\author{Rogue Intelligence LNC.\\Valhalla Unified Core (VUC)}
\date{2026-06-27}

\begin{document}
\maketitle

\begin{abstract}
Large language models conflate \emph{memory}, \emph{routing}, and \emph{generation} inside a single weight matrix, forcing domain adaptation via fine-tuning or ever-growing prompts. We present \textbf{ValhallaBase}, an inference-time paradigm that incubates persistent \textbf{structure memory} (Hub, Tile, StemCell, Fate) \emph{before} optional language-model decoding. Structure is updated by zero-weight session ingest---no SGD, LoRA, or corpus dump into the prompt window. A \textbf{hybrid router} sends open questions to native structure retrieval and multiple-choice items to LM logprob heads; every answer exports auditable routing metadata. An \textbf{algebraic Fate Output Gate (FOG)} and \textbf{proactive idle} probes decide \emph{when} to speak, not \emph{what} to say. Cross-language queries hit the same canonical corpus line without model switching.

On pre-registered batteries with Wilson 95\% confidence intervals: fair holdout RAG accuracy \textbf{96.8\%} [83.8\%, 99.4\%] (30/31), statistically tied with Qwen2.5-0.5B RAG; \textbf{transfer fresh retrieval} rescues polluted sessions by \textbf{+61.3\,pp} [+45.2, +77.4]; FOG emitted precision \textbf{96.6\%}; proactive on-topic precision given emit \textbf{100\%}; 30-turn topic continuity \textbf{96.7\%}; external holdout and local corpus demos \textbf{100\%}. Context append carries $\sim$64 characters per wire step versus $\sim$752-character full-prompt resends ($3.62\times$ growth) on a 12-step growth curve---structure changes \emph{how} context is carried, not merely RAG ceiling accuracy.
\end{abstract}

\section{Introduction}

Vertical AI deployments require sovereignty, auditability, and domain structure that survives beyond one context window. Benchmark leaderboard models optimize fluency on static splits; enterprises need:

\begin{enumerate}
  \item \textbf{Persistent structure} as a first-class artifact, not ephemeral prompt text.
  \item \textbf{Honest typing}: open, numeric, and MCQ items fail differently.
  \item \textbf{Session protocols} that avoid retrieval pollution when corpora grow.
  \item \textbf{Output governance}: abstain, defer, or proactive nudge with traceable gates.
  \item \textbf{Cost transparency}: context transport and LM prefill tokens, not accuracy alone.
\end{enumerate}

ValhallaBase addresses these by separating \textbf{structure incubation} from \textbf{decode}. A compact instruction-tuned LM (Qwen2.5-0.5B-Instruct) acts as an optional patch head; the default open path uses native corpus memory over a TriadIncubationSnapshot.

\paragraph{Distinctive claims (not generic RAG).}
\begin{itemize}
  \item \textbf{D1 Structure-first:} incubate, then decode---LM reads a patched snapshot.
  \item \textbf{D2 Zero-weight session learning:} targeted ingest without gradient updates.
  \item \textbf{D3 Hybrid type routing:} open $\to$ native; MCQ $\to$ lm\_patch logprob.
  \item \textbf{D4 Session routing discipline:} fresh per-question retrieval vs.\ polluted dump (+61\,pp lift).
  \item \textbf{D5 Context-flexible append:} $\sim$64\,chars/step wire payload vs.\ full prompt resend.
  \item \textbf{D6 Follow-up-aware retrieval:} question history boosts in \texttt{native\_qa.rs}.
  \item \textbf{D7 FOG algebraic gate:} E7-integer $\tau=12/(138+k)$; emit/abstain/defer.
  \item \textbf{D8 Proactive Fate:} idle self-probe with overlap gate; 100\% on-topic given emit.
  \item \textbf{D9 Cross-language unified core:} EN/ZH/FR $\to$ same \texttt{[corpus:0]} line.
  \item \textbf{D10 Non-isolated multimodal:} interleaved session pulses vs.\ late fusion ($0.964$ vs $0.943$ universal score).
  \item \textbf{D11 Auditable JSON routing:} \texttt{decode\_path}, FOG decisions per answer.
  \item \textbf{D12 E7 coordination:} spread gates derived from algebra, not fitted floats.
\end{itemize}

\section{Related Work}

\textbf{RAG} retrieves flat passages into the prompt; Valhalla retrieves over an incubated structure checkpoint with optional LM patch, not text blocks alone.

\textbf{Adapter / LoRA fine-tuning} updates weights; Valhalla uses zero-weight ingest with explicit falsification flags when holdout lift equals targeted RAG.

\textbf{Memory-augmented networks} learn memory end-to-end; Valhalla uses rule- and score-based native QA plus Fate export for audit.

\textbf{Mixture-of-experts} routes internal layers; Valhalla routes at product level across native memory, structure MCQ, and patched LM with compliance-grade traces.

\textbf{Hybrid structure--LM systems} (cf.\ our fair-1.2 memory paper) established 80.69\% on 145 typed prompts; this work extends to inference-paradigm batteries, confidence intervals, FOG, proactive idle, cross-language core unity, and context transport metrics.

\section{Method}

\subsection{Triad structure substrate}

Text and multimodal ingress encode into body ports. Hub, Tile, StemCell, and Fate run under \texttt{quad\_cycle} ingress routing. Incubation yields a \textbf{TriadIncubationSnapshot} forked per question. Optional \texttt{fate\_qa\_feedback} closes a quality loop without LM gradients.

\paragraph{Stem organ clusters and Tile aggregation (measured).}
Stem cells form \texttt{stem\_organ\_clusters} via connected components on the cell graph; signatures enter \texttt{score\_structure\_alignment} (organ weight $0.12$). Tiles merge/split into \texttt{completed\_tiles} (weight $0.15$). On fair-1.2 corpus ($\sim$49 ingest lines), we observe $\sim$11 completed tiles and typically \textbf{one} stem mega-cluster ($\sim$159 cells)---structural aggregation runs, but functional organ differentiation is not yet evidenced. Persistent incubation lifts native MCQ (tile body 47.8\% vs.\ 28.3\% isolated; +19.6\,pp) while production MCQ remains lm\_patch-dominated (65.2\% vs.\ structure-only 30.4\%). Analysis: \texttt{STEM\_TILE\_ORGAN\_AGGREGATION\_ANALYSIS\_20260628.md}.

\paragraph{Backbone universality (measured).}
Protocol \texttt{backbone-universality-v1} runs the \textbf{same} Valhalla hybrid structure stack once while swapping trad\_lm RAG decode across Qwen2.5-0.5B, Gemma-2-2b-it, and Phi-3-mini-4k-instruct. External holdout: structure hybrid \textbf{15/15 (100\%)}. Open\_retrieval slice ($n=31$, per-question \texttt{corpus\_for\_prompt}): structure hybrid \textbf{28/31 (90.3\%)}; all three trad\_lm arms \textbf{31/31}. Fair 61-question open mix (includes open\_generation): structure 45.9\% vs.\ trad\_lm up to 98.4\%---not comparable to holdout transfer protocol. Rust \texttt{lm\_patch} remains Qwen2-only and is outside this battery. Summary: \texttt{BACKBONE\_UNIVERSALITY\_SUMMARY\_20260628.md}.

\subsection{Zero-weight session ingest}

Corpus lines merge via \texttt{merge\_context} or \textbf{targeted} top-$k$ per question---never full dump before batch eval. Production eval protocol:
\begin{enumerate}
  \item ingest targeted lines per question;
  \item optional curriculum QA;
  \item eval: clear corpus $\to$ per-question targeted append;
  \item forbid batch eval on 80+ line polluted sessions.
\end{enumerate}

\subsection{Hybrid decode routing}

Production stack:
\begin{verbatim}
ValhallaBase(
  decode="hybrid",
  follow_up_decode="native_follow_up_aware",
  fate_ingress_routing="quad_cycle",
  fate_qa_feedback=True,
)
\end{verbatim}
Open retrieval uses native structure memory; MCQ uses LM logprob on patched weights; numeric uses $\max$(structure, patch).

\subsection{Fate Output Gate (FOG v2)}

Emit iff StructureReady $\wedge$ EvidenceReady $\wedge$ TriggerPresent. Version~2 binds thresholds to E7 integers: $\tau = 12/(138+k)$, margin $\Delta \ge s_1/28$. FOG decides \emph{permission to speak}; native QA still selects content.

\subsection{Proactive idle}

On idle timeout, Fate builds a self-probe from max token overlap with prior turns (CJK bigram support), passes FOG, and may emit a grounded nudge. Mismatch corpora trigger abstain (MAR) rather than hallucinated interrupt (FIR).

\subsection{Context transport (token-relevant)}

Internal context grows with session appends; wire JSON carries incremental line payloads ($\sim$64.3 chars average per step in growth-curve protocol). Traditional LM baselines resend the full corpus block each turn ($3.62\times$ character growth over 12 steps). \textbf{LM prefill token totals} are planned under protocol \texttt{token-efficiency-v1} (Qwen tokenizer, 30-turn matched accuracy).

\section{Experimental Setup}

\begin{table}[h]
\centering
\small
\begin{tabular}{ll}
\toprule
Setting & Value \\
\midrule
Companion LM & Qwen2.5-0.5B-Instruct (+ optional patch 0.08) \\
Fair holdout & open\_retrieval slice, $n=31$ \\
Confidence & Wilson 95\% (proportions); bootstrap 95\% (paired lifts) \\
Aggregated report & \texttt{confidence-report-v1} \\
Corpus demo & 20 probes, \texttt{valhalla-unified-core/docs} \\
External holdout & 15 cross-domain probes \\
Cross-language & EN/ZH/FR, \texttt{crosslang-core-unified-v1} \\
\bottomrule
\end{tabular}
\end{table}

\section{Results}

\subsection{Grounding and session routing}

\begin{table}[h]
\centering
\begin{tabular}{lc}
\toprule
Metric & Estimate [95\% CI] \\
\midrule
Fair holdout cold RAG & 96.8\% [83.8\%, 99.4\%] (30/31) \\
Polluted session & 35.5\% [21.1\%, 53.1\%] \\
Transfer fresh & 96.8\% [83.8\%, 99.4\%] \\
Transfer vs.\ polluted lift & +61.3\,pp [+45.2, +77.4] \\
Traditional LM RAG & 96.8\% [83.8\%, 99.4\%] \\
Valhalla $-$ trad\_lm & +0.0\,pp [$-$9.7, +9.7] \\
TPI-v2 index & 0.790 [0.740, 0.823] \\
NPPI index & 0.884 [0.848, 0.984] \\
\bottomrule
\end{tabular}
\caption{Structure matches RAG accuracy on holdout; session routing is the differentiator.}
\end{table}

\subsection{Output governance and dialogue}

\begin{table}[h]
\centering
\begin{tabular}{lc}
\toprule
Metric & Estimate [95\% CI] \\
\midrule
FOG emitted precision & 96.6\% [82.8\%, 99.4\%] \\
Proactive UP $|$ emit & 100\% [87.5\%, 100\%] \\
Proactive MAR (mismatch abstain) & 88.2\% [73.4\%, 95.3\%] \\
30-turn on-topic & 96.7\% [83.3\%, 99.4\%] \\
Swap fidelity & 100\% [75.8\%, 100\%] \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-language and external generalization}

Cross-language core unified: CCU $=0.950$, 6/6 queries hit identical corpus text. External holdout: 15/15 (100\%) [79.6\%, 100\%]. Local corpus demo: 20/20 (100\%) [83.9\%, 100\%].

\subsection{MCQ coverage and deploy routing}

Protocol \texttt{mcq-coverage-v1} ($n=46$ fair-1.2 MCQ): production hybrid lm\_patch \textbf{65.2\%} [50.8\%, 77.3\%]; traditional LM RAG \textbf{28.3\%}; oracle max(native, patch) \textbf{69.6\%} (not deployable). Native \texttt{mcq\_option} alone remains \textbf{30.4\%} --- MCQ requires lm\_patch or deploy \texttt{mcq\_decode=max\_pick} (native + patch pick by margin, Rust \texttt{ValhallaMcqDecode::MaxPick}).

\subsection{Context transport and token efficiency}

On \texttt{context-flexibility-max-goal-v1} (12 append steps): average incremental payload $\sim$64.3 characters; traditional full prompt 752 characters ($3.62\times$ seed). Internal snapshot: 638 characters / 16 lines.

Protocol \texttt{token-efficiency-v1} compares matched 30-turn threads (3 topics $\times$ 10 turns) using the Qwen2.5-0.5B tokenizer: Valhalla session append vs.\ RAG full resend vs.\ RAG with dialogue history. Table~\ref{tab:token-efficiency} unifies wire cost, LM prefill tokens, decode tokens, and accuracy---substantiating infrastructure value at matched task quality.

\begin{table}[h]
\centering
\small
\caption{Wire + token + accuracy at matched 30-turn dialogue (token-efficiency-v1). Holdout accuracy from fair open\_retrieval holdout ($n=31$).}
\label{tab:token-efficiency}
\begin{tabular}{lrrrr}
\toprule
Architecture & Wire chars & Prefill tokens & Decode tokens & Accuracy \\
\midrule
Valhalla append (native) & 2{,}319 & 1{,}126 & 0 & 96.8\% holdout \\
RAG full resend & 14{,}268 & 3{,}868 & 3{,}840 & 96.8\% holdout \\
RAG + history & --- & 11{,}845 & 3{,}840 & 96.8\% holdout \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Interpretation:} Valhalla wire ratio vs.\ RAG full resend is $0.16\times$; prefill ratio is $0.29\times$ (question-only vs.\ full corpus each turn). RAG-with-history prefill reaches 11{,}845 tokens over 30 turns. Accuracy remains statistically tied on holdout---efficiency is not purchased by sacrificing grounding.

\section{Discussion}

\paragraph{Where structure wins.}
Session routing (+61\,pp), FOG/proactive governance, cross-language core unity, multimodal non-isolated alignment, and flat transport cost---not raw open\_retrieval ceiling versus RAG LM (statistical tie at 96.8\%).

\paragraph{Honest falsification flags.}
\texttt{SESSION\_SNAPSHOT\_NO\_HOLDOUT\_LIFT\_BEYOND\_RAG}; \texttt{NOT\_WEIGHT\_FINE\_TUNE\_REPLACEMENT}; \texttt{STEM\_ORGAN\_NOT\_FUNCTIONALLY\_DIFFERENTIATED}; \texttt{MCQ\_STRUCTURE\_BELOW\_LM\_PATCH}; \texttt{LM\_PATCH\_QWEN2\_ONLY}; \texttt{BACKBONE\_BATTERY\_PROTOCOL\_DIFFERS\_FROM\_TRANSFER}. These pre-registered limits strengthen rather than weaken the paradigm claim.

\paragraph{Token metrics as evidence.}
We recommend reporting three layers: (1)~wire bytes/chars per session step; (2)~LM prefill tokens cumulative vs.\ RAG baseline at matched accuracy; (3)~decode tokens by path (native vs.\ patch). Accuracy alone cannot substantiate infrastructure value.

\paragraph{Limitations.}
Single small LM; fair holdout $n=31$; proactive pass rate 63/72 under strict overlap gates; token-efficiency-v1 pending.

\section{Conclusion}

ValhallaBase is a reproducible \textbf{inference-time AI paradigm}: structure-first memory, zero-weight ingest, hybrid auditable routing, algebraic output gating, and context-efficient sessions. It matches compact LM RAG on holdout accuracy while offering qualitatively different capabilities---pollution-safe routing, proactive grounded dialogue, and cross-language unified cores---that prompt-only systems cannot audit or replicate without protocol changes.

\paragraph{Reproducibility.}
Scripts under \texttt{tools/valhalla\_inference/}; reports in \texttt{reports/valhalla\_inference/}; companion doc \texttt{docs/DISTINCTIVE\_CAPABILITIES\_CN.md}.

\begin{thebibliography}{9}
\bibitem{hybrid2026}
Rogue Intelligence LNC.,
\textit{Hybrid Structure Memory with Language-Model Patch Routing},
fair-1.2 benchmark, 2026.
\bibitem{confidence2026}
\textit{Valhalla Inference Confidence Report v1.1},
protocol \texttt{confidence-report-v1}, 2026-06-27.
\bibitem{contextflex2026}
\textit{Context Flexibility Maximum Goal},
protocol \texttt{context-flexibility-max-goal-v1}, 2026-06-25.
\end{thebibliography}

\end{document}
